### 2022-06-13 paper.tex
add workflow fig caption

\begin{figure*}
    \begin{center}
        \includegraphics[width=.85\textwidth]{figures/Fig_workflow.png}
    \end{center}
    \caption{\small \textbf{Overview of the functionalities implemented in the scbs package}
    }
    \label{workflow}
\end{figure*}

We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{workflow}). The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark} or methylpy \citep{methylpy}. For each cell, these tools produce a text file that lists the methylation status of all CpG sites. Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.

In brief, we store methylation of each chromosome in a matrix where each column represents a cell and each row represents a base pair. Methylated and unmethylated sites are coded as $+1$ and $-1$, respectivelty. Since the vast majority of values in this matrix are missing due to the sparsity of scBS data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format. This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.

>>>>

\begin{figure*}
    \begin{center}
        \includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
    \end{center}
    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}
    \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimised for efficient access to genomic intervals.
    \texttt{scbs} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
    These cells can be discarded with \texttt{scbs filter}.
    To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
    The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably-methylated regions) in the data with \texttt{scbs scan}.
    The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
    }
    \label{workflow}
\end{figure*}

We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{workflow}). The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark} or methylpy \citep{methylpy}. For each cell, these tools produce a text file that lists the methylation status of all CpG sites. Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome. \texttt{scbs prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or even custom user-defined formats.

In brief, we store methylation of each chromosome in a matrix where each column represents a cell and each row represents a base pair. Methylated and unmethylated sites are coded as $+1$ and $-1$, respectively. Since the vast majority of values in this matrix are missing due to the sparsity of scBS data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format. This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.
